\section{Introduction}
\label{sec:intro}

Object detection systems have achieved remarkable performance on standard benchmarks; however, their reliability degrades significantly in the presence of out-of-distribution (OoD) inputs. A critical manifestation of this issue is hallucinated detections, where models assign high confidence to non-existent or irrelevant objects. This poses serious concerns in real-world deployments such as autonomous driving, surveillance, and medical imaging.

Recent work, including \textit{Revisiting Out-of-Distribution Detection in Real-time Object Detection: From Benchmark Pitfalls to a New Mitigation Paradigm}, has highlighted limitations in existing OoD detection benchmarks and emphasized the need for robust mitigation strategies. Most prior approaches rely on confidence re-scoring, energy-based methods, or training with curated OoD datasets. While effective to some extent, these methods often ignore an important factor: the intrinsic signal properties of input regions that trigger false detections.

Deep neural networks are known to exhibit spectral bias, favoring certain frequency components in input data. In the context of object detection, this leads to an over-reliance on high-frequency textures and fine-grained patterns, which may not correspond to semantically meaningful objects. Furthermore, small bounding boxes often lack sufficient contextual information, making them more prone to misclassification and inflated objectness scores.

In this work, we address hallucinated detections in OoD settings through a combination of post-hoc calibration and explainability-driven analysis. Rather than relying on architectural changes or retraining strategies, we focus on lightweight inference-time corrections and behavioral diagnostics of detection models.

We propose a frequency-aware calibration mechanism that operates directly on detection outputs. This method suppresses spurious detections by penalizing high-frequency regions using Laplacian-based energy estimation, combined with size-aware suppression for small bounding boxes. The resulting correction is applied post-hoc to predicted confidence scores, making the method training-free, model-agnostic, and compatible with real-time YOLO-based detectors.

To better understand model behavior under distribution shift, we complement our calibration approach with explainability techniques. We use Grad-CAM to visualize spatial regions influencing detector predictions, and occlusion sensitivity analysis to evaluate how local perturbations affect confidence and localization stability.

To validate the effectiveness of the proposed framework, we conduct experiments on both in-distribution (Pascal VOC) and OoD datasets constructed from Caltech-256. We evaluate performance using the following metrics and analysis tools:

\begin{itemize}
    \item Hallucination Rate (HR) and False Positive (FP) reduction under OoD settings
    \item Confidence distribution shifts between baseline and calibrated models
    \item Intersection-over-Union (IoU) based localization consistency
    \item Pointing Game metric for spatial alignment evaluation
    \item Insertion and Deletion curves for explanation faithfulness
\end{itemize}

Our results demonstrate that frequency-aware calibration consistently reduces hallucinated detections while improving interpretability of model behavior under distribution shift. The combination of lightweight post-hoc correction and explainability analysis provides both practical improvements and deeper insights into failure modes of object detectors in OoD environments.